Categorical Framework for Visualization Software Design

### 2025-11-27 paper/main.tex
homemeomorphism

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}, but generally do not verify that input structure. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner because it introduces ambiguity over which part of the line maps back to the data. A map that by definition has continuous maps between the input and output spaces, such as data and graphics, is called a \textit{homeomorphism}\cite{riehlCategoryTheoryContext}. The bar plot, line plot, and heatmap in \autoref{fig:related-work:continuity:ktypes} have a homeomorphic relationship to the 0D ($\bullet$) points, 1D (--) linear, and 2D ($\blacksquare$) surface continuities embedded in the continuous 3 dimensional surface encapsulating time and position because each point of the visualization maps back into a point in its corresponding indexing space in the cube. Using homeomorphism to test whether continuity is preserved formalizes Bertin's codification of how the topology of observations matches the class of representation (i.e. point , line, area) \cite{bertinSemiologyGraphicsDiagrams2011} and Wilkinson's assertion that connectivity must be preserved \cite{wilkinsonGrammarGraphics2005}. Unlike Wilkinson's

>>>>

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}, but generally do not verify that input structure. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner because it introduces ambiguity over which part of the line maps back to the data. This ambiguity cannot exist when the map between data and graphic is a homeomorphism
\begin{definition}
  A function $f$ is a \textit{homeomorphism} if it is bijective, continuous, and has a continuous inverse function $f^{-1}$.
\end{definition}
which means that every element of the graphic maps back to a data element. The bar plot, line plot, and heatmap in \autoref{fig:related-work:continuity:ktypes} have a homeomorphic relationship to the 0D ($\bullet$) points, 1D (--) linear, and 2D ($\blacksquare$) surface continuities embedded in the continuous 3 dimensional surface encapsulating time and position because each point of the visualization maps back into a point in its corresponding indexing space in the cube. Using homeomorphism to test whether continuity is preserved formalizes Bertin's codification of how the topology of observations matches the class of representation (i.e. point , line, area) \cite{bertinSemiologyGraphicsDiagrams2011} and Wilkinson's assertion that connectivity must be preserved \cite{wilkinsonGrammarGraphics2005}. Unlike Wilkinson's
